---
title: "Introduction to Data Science"
subtitle: "Session 4: Relational databases"
author: "Simon Munzert"
institute: "Hertie School | [GRAD-C11/E1339](https://github.com/Fall-2021-Introduction-to-Data-Science)" #"`r format(Sys.time(), '%d %B %Y')`"
output:
  xaringan::moon_reader:
    css: [default, '../simons-touch.css', metropolis, metropolis-fonts] 
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'
      hash: true
---


```{css, echo=FALSE} 
@media print { # print out incremental slides; see https://stackoverflow.com/questions/56373198/get-xaringan-incremental-animations-to-print-to-pdf/56374619#56374619
  .has-continuation {
    display: block !important;
  }
}
```

```{r setup, include=FALSE}
# figures formatting setup
options(htmltools.dir.version = FALSE)
library(knitr)
opts_chunk$set(
  prompt = T,
  fig.align="center", #fig.width=6, fig.height=4.5, 
  # out.width="748px", #out.length="520.75px",
  dpi=300, #fig.path='Figs/',
  cache=T, #echo=F, warning=F, message=F
  engine.opts = list(bash = "-l")
  )

## Next hook based on this SO answer: https://stackoverflow.com/a/39025054
knit_hooks$set(
  prompt = function(before, options, envir) {
    options(
      prompt = if (options$engine %in% c('sh','bash')) '$ ' else 'R> ',
      continue = if (options$engine %in% c('sh','bash')) '$ ' else '+ '
      )
})
```


# Table of contents

1. [Tidyverse basics](#basics)<sup>1</sup>

2. [Pipes](#pipes)

3. [Data wrangling with dplyr](#dplyr)

4. [Data tidying with tidyr](#tidyr)

5. [Tidy programming](#programming)

6. [Coding style](#style)

7. [Summary](#summary)

.footnote[<sup>1</sup> Parts of this lecture draws on materials from Grant McDermott's excellent [*Data Science for Economists*](https://github.com/uo-ec607/lectures) class.]


---
name: joins

# Joins

One of the mainstays of the dplyr package is merging data with the family [join operations](https://cran.r-project.org/web/packages/dplyr/vignettes/two-table.html).
- `inner_join(df1, df2)`
- `left_join(df1, df2)`
- `right_join(df1, df2)`
- `full_join(df1, df2)`
- `semi_join(df1, df2)`
- `anti_join(df1, df2)`

(You find find it helpful to to see visual depictions of the different join operations [here](https://r4ds.had.co.nz/relational-data.html).)

--

For the simple examples that I'm going to show here, we'll need some data sets that come bundled with the [**nycflights13**](http://github.com/hadley/nycflights13) package. 
- Load it now and then inspect these data frames in your own console.

```{r flights, echo = F}
library(nycflights13)
```
```{r, eval = F}
library(nycflights13)
flights 
planes
```

---

# Joins *cont.*

Let's perform a [left join](https://stat545.com/bit001_dplyr-cheatsheet.html#left_joinsuperheroes-publishers) on the flights and planes datasets. 
- *Note*: I'm going subset columns after the join, but only to keep text on the slide.

--

```{r join1}
left_join(flights, planes) %>%
  select(year, month, day, dep_time, arr_time, carrier, flight, tailnum, type, model)
```

---

# Joins *cont.*

(*continued from previous slide*)

Note that dplyr made a reasonable guess about which columns to join on (i.e. columns that share the same name). It also told us its choices: 

```
*## Joining, by = c("year", "tailnum")
```

However, there's an obvious problem here: the variable "year" does not have a consistent meaning across our joining datasets!
- In one it refers to the *year of flight*, in the other it refers to *year of construction*.

--

Luckily, there's an easy way to avoid this problem. 
- See if you can figure it out before turning to the next slide.
- Try `?dplyr::join`.

---

# Joins *cont.*

(*continued from previous slide*)

You just need to be more explicit in your join call by using the `by = ` argument.
- You can also rename any ambiguous columns to avoid confusion. 
```{r join2}
left_join(
  flights,
  planes %>% rename(year_built = year), ## Not necessary w/ below line, but helpful
  by = "tailnum" ## Be specific about the joining column
  ) %>%
  select(year, month, day, dep_time, arr_time, carrier, flight, tailnum, year_built, type, model) %>%
  head(3) ## Just to save vertical space on the slide
```

---

# Joins *cont.*

(*continued from previous slide*)

Last thing I'll mention for now; note what happens if we again specify the join column... but don't rename the ambiguous "year" column in at least one of the given data frames.
```{r join3}
left_join(
  flights,
  planes, ## Not renaming "year" to "year_built" this time
  by = "tailnum"
  ) %>%
  select(contains("year"), month, day, dep_time, arr_time, carrier, flight, tailnum, type, model) %>%
  head(3)
```

--

Make sure you know what "year.x" and "year.y" are. Again, it pays to be specific.




